---
title: "Figures and analysis to accompany 'Canine Sniffing Biomechanic Responses to Varied Physical Properties'"
author: 
- Lydia Burnett^1^, Lindsay Waldrop^2^, Nick Hebdon^2^, Pete Stevens^3^,  
- Monica Moljo^1^, Lauryn DeGreeff^1^
date: "`r Sys.Date()`"
output: html_notebook
---
```{r setup}
knitr::opts_chunk$set(echo=FALSE)
```

# Instructions for running code

This code notebook contains all code needed to reproduce the figures and statistics of the manuscript. It was run with R version 4.4.0 (2024-04-24) and RStudio v 2024.04.1+748 (2024.04.1+748). 

To rerun the code in this file: 

 1. Clone the Github project at https://github.com/lindsaywaldrop/dog-sniffs
 2. Open RStudio and the RProject file dog-sniffs.Rproj. 
 3. Open the Reproduction.Rmd and click Run and Run all. 

Most code is not echoed in the HTML notebook file, but it is present in the RMD file. 

Note: the terminology in the code often refers to sport dogs as "civilian" and law-enforcement dogs as "operational". 

# Reproduction of full analysis

Coming Soon!

# Reproduction of data-driven figures in manuscript

This RMD will call a csv file in the results directory produced from a previous run of the full analysis code from raw sensor data with the date: 

```{r analysis-date, echo=TRUE}
analysis_date <- "2023-08-30"
```

If you would like to rerun the full analysis, please change the above date to the date which you ran the full analysis code. 

Loads required libraries: 

```{r libraries, echo=TRUE}
library(knitr)
library(ggrepel)
library(here)
library(ggplot2)
library(viridis)
library(patchwork)
library(forcats)
library(dplyr)
library(signal)
library(pracma)
```


Loading results and performing initial calculations:

```{r loading-results, message=FALSE, warning=FALSE}
# Loading custom functions (located in src/ folder)
source("./src/data_handling_fxns.R")
source("./src/ethogram_fxns.R")
source("./src/kinematic_fxns.R")

frame_rate <- 29.97 # Setting frame rate of videos

# Loading the results of previous analysis
sniff_results <- load_sniff_results(analysis_date)

# Loading snout lengths of sport dogs
snout_lengths <- calculate_snout_lengths(sniff_results)

# Loading dog weights
dog_sizes <- read.csv("./data/kinematics/dog_sizes.csv")

# Assigning snout lengths, weights, and size classifications to main data set
sniff_results$snout_length <- rep(NA, nrow(sniff_results))
sniff_results$size <- rep(NA, nrow(sniff_results))
sniff_results$weight <- rep(NA, nrow(sniff_results))
for(j in 1:nrow(sniff_results)){
  sniff_results$snout_length[j] <- snout_lengths$head_mean[snout_lengths$dog == 
                                                             sniff_results$dog[j]]
  sniff_results$size[j] <- dog_sizes$size[dog_sizes$dog == 
                                            sniff_results$dog[j]]
  sniff_results$weight[j] <- dog_sizes$weight[dog_sizes$dog == 
                                            sniff_results$dog[j]]
}
sniff_results$weight <- sniff_results$weight/2.205 # Converting lbs to kg

# Adding information on the training status for dogs (sport dogs were trained in trial 3, trials 4 and 5 were law-enforcement dogs)
sniff_results$trained <- ifelse(sniff_results$trial == 3 | sniff_results$trial == 4, T, F)

# Normalizing volumes and volume flow rates with body weight
sniff_results$norm_volume <- sniff_results$sniff_volume/sniff_results$weight
sniff_results$norm_total_volume <- sniff_results$total_volume/sniff_results$weight

# Removing dog 102
sniff_results <- sniff_results[sniff_results$dog != 102, ]

# Reordering results 
sniff_this <- sniff_results %>% 
  mutate(dog = fct_reorder(dog, -as.numeric(as.character(snout_length))))
# Factoring size data
sniff_this$size <- factor(sniff_this$size, ordered = TRUE, levels = c("small", "large", "operational"))

# Summarize results by dog number, size, weight, snout length, and trial
sniff_summaries <- sniff_this %>%
  group_by(dog, size, snout_length, trial, weight) %>%
  summarise(sniff_vfr = mean(sniff_vfr, na.rm = T),                   # Mean volume flow rate of sniffs in a burst
            sniff_volume = mean(sniff_volume, na.rm = T),             # Mean sniff volume
            norm_volume = mean(norm_volume, na.rm = T),               # Mean sniff volume normalized by body weight
            sniff_freq = mean(sniff_freq, na.rm = T),                 # Mean sniff frequency in Hz
            sniff_duration = mean(sniff_duration, na.rm = T),         # Mean sniff duration in seconds
            sniff_counts = mean(sniff_counts, na.rm = T),             # Total count of sniffs in a burst
            total_volume = mean(total_volume, na.rm = T),             # Mean total sampling volume of sniffs in a burst
            norm_total_volume = mean(norm_total_volume, na.rm = T))   # Mean total sampling volume of sniffs in a burst normalized by body weight

# Summarize results by dog number, size, weight, snout length, and trial
sniff_summaries_chemical <- sniff_this %>%
  group_by(dog, size, snout_length, trial, weight, chemical) %>%
  summarise(sniff_vfr = mean(sniff_vfr, na.rm = T),                   # Mean volume flow rate of sniffs in a burst
            sniff_volume = mean(sniff_volume, na.rm = T),             # Mean sniff volume
            norm_volume = mean(norm_volume, na.rm = T),               # Mean sniff volume normalized by body weight
            sniff_freq = mean(sniff_freq, na.rm = T),                 # Mean sniff frequency in Hz
            sniff_duration = mean(sniff_duration, na.rm = T),         # Mean sniff duration in seconds
            sniff_counts = mean(sniff_counts, na.rm = T),             # Total count of sniffs in a burst
            total_volume = mean(total_volume, na.rm = T),             # Mean total sampling volume of sniffs in a burst
            norm_total_volume = mean(norm_total_volume, na.rm = T))   # Mean total sampling volume of sniffs in a burst normalized by body weight

# Assigning target column to distinguish between target (2E1H or ammonia) or non-target odors
sniff_summaries_chemical$target <- ifelse(sniff_summaries_chemical$chemical == "2E1H", "yes",
                                          ifelse(sniff_summaries_chemical$chemical == "Ammonia", "yes", "no"))

# Summarize results by dog number, size, snout length and weight, this summarizes data across trials (for only events that result in positive volumes)
sniff_summaries_alltrials <- sniff_this[sniff_this$sniff_volume > 0, ] %>%
  group_by(dog, size, snout_length, weight) %>%
  summarise(sniff_vfr = mean(sniff_vfr, na.rm = T),                   # Mean volume flow rate of sniffs in a burst
            sniff_volume = mean(sniff_volume, na.rm = T),             # Mean sniff volume
            norm_volume = mean(norm_volume, na.rm = T),               # Mean sniff volume normalized by body weight
            sniff_freq = mean(sniff_freq, na.rm = T),                 # Mean sniff frequency in Hz
            sniff_duration = mean(sniff_duration, na.rm = T),         # Mean sniff duration in seconds
            sniff_counts = mean(sniff_counts, na.rm = T),             # Total count of sniffs in a burst
            total_volume = mean(total_volume, na.rm = T),             # Mean total sampling volume of sniffs in a burst
            norm_total_volume = mean(norm_total_volume, na.rm = T))   # Mean total sampling volume of sniffs in a burst normalized by body weight

# Remove extra NA line at the end of the data set:
sniff_summaries_alltrials <- sniff_summaries_alltrials[!is.na(sniff_summaries_alltrials$dog),]

# Assigning operational status to each dog (operational = law-enforcement, civilian = sport)
sniff_summaries_alltrials$op <- ifelse(sniff_summaries_alltrials$size == "operational",
                                       "operational", "civilian")


```


Setting color palettes and labels for figures:

```{r fig-preferences, include=FALSE}
trial_cols <- c("deepskyblue4","cadetblue1", "lightsalmon",  "gold")
sniff_this <- sniff_this[order(sniff_this$weight),]
sniff_this$dog <- factor(sniff_this$dog, ordered = T, levels = unique(sniff_this$dog))
chem_labels <- c("2E1H\n(target)", "Ammonia\n(target)", "Blank", 
                              "1-BO\n(non-target)", 
                              "MB\n(non-target)")
```

## Figure 2

Code loads in raw sensor data (in the data/flowsensor directory), filters, and clips to a snapshot to produce figure 2. 

```{r sniff-trace-calcs}
## Loading data for Figure 2

sample_rate <- 100  # sampling rate in samples per second
N <- 4              # Butterworth filter polynomial fit coefficient
filsig <- 0.15      # Butterworth filter signal frequency for low pass 
sniff_parameters <- load_sniff_parameters() # Loading the parameter data set

# Loading data for trial 4, dog 100, run 2, bromo-octane
dat_a_full <- load_data(trial = 4, 
                     dog = 100,
                     run = 2,
                     chemical = "bromooctane",
                     sample_rate, 
                     min_stop = sniff_parameters$cal[722], 
                     F)

# Filtering data with forward-reverse low-pass Butterworth filter
dat_a_full <- filter_data(dat_a_full, F, F, N, filsig)

# Restricting data set to snapshot (Fig 2B)
dat_a_snap <- dat_a_full[dat_a_full$adj_time >= sniff_parameters$start_time[722] & 
                           dat_a_full$adj_time <= sniff_parameters$end_time[722],]

# Loading data for trial 1, dog 11, run 1, bromo-octane
dat_c_sniff <- load_data(trial = 1, 
                     dog = 11,
                     run = 1,
                     chemical = "bromooctane",
                     sample_rate, 
                     min_stop = sniff_parameters$cal[29], 
                     F)
# Filtering data with forward-reverse low-pass Butterworth filter
dat_c_sniff <- filter_data(dat_c_sniff, T, F, N, filsig)

# Restricting data set to snapshot (Fig 2C)
dat_c_sniff <- dat_c_sniff[dat_c_sniff$adj_time >= 73.2 & 
                        dat_c_sniff$adj_time <= 74,]

# Loading data for trial 4, dog 100, run 1, blank
dat_d_snort <- load_data(trial = 4, 
                     dog = 100,
                     run = 1,
                     chemical = "Blank",
                     sample_rate, 
                     min_stop = sniff_parameters$cal[719], 
                     F)

# Filtering data with forward-reverse low-pass Butterworth filter
dat_d_snort <- filter_data(dat_d_snort, F, F, N, filsig)

# Restricting data set to snapspot (Fig 2D)
dat_d_snort <- dat_d_snort[dat_d_snort$adj_time >= 163.2 & 
                        dat_d_snort$adj_time <= 164,]
```

Figure 2 reproduction:

```{r figure-2, echo=FALSE, warning=FALSE, fig.cap="Example of sniffing data as generated by the sensors. A: complete trace of Trial 4, Dog 100, Run 2; sniff cycle of 1-BO, gray box indicates focus of part B. B: shows a typical sniff cycle pattern within A. C: Trial 1, Dog 11, Run 1, sniff of the blank tube, showing isolated individual sniffs. D: Trial 4, Dog 100, Run 1, sniff of the blank tube showing a snort. Black lines are unfiltered data and red lines are filtered data."}
# Figure 2

box <- data.frame("x" = c(62, 59.1, 59.1, 62), "y" = c(-10, -10, 18, 18))

p1 <- ggplot(dat_a_full[dat_a_full$adj_time > 25,], aes(adj_time, sample)) + 
  geom_line() +
  geom_polygon(data = box, aes(x,y), col = "gray40", fill = NA, type = 2) +
  xlab("Time (s)") + ylab("Volume flow (std. L/min)") + 
  theme_bw()

p2 <- ggplot(dat_a_snap, 
       aes(adj_time, sample)) + 
  geom_line()  +
  geom_line(aes(adj_time, sample_filtered), col = "red") + 
  xlab("Time (s)") + ylab("Volume flow (std. L/min)") + 
  theme_bw()

p3 <- ggplot(dat_c_sniff, 
       aes(adj_time, sample)) + 
  geom_line()  +
  geom_line(aes(adj_time, sample_filtered), col = "red") + 
  xlab("Time (s)") + ylab("Volume flow (std. L/min)") + 
  theme_bw()

p4 <- ggplot(dat_d_snort, 
       aes(adj_time, sample)) + 
  geom_line()  +
  geom_line(aes(adj_time, sample_filtered), col = "red") + 
  xlab("Time (s)") + ylab("Volume flow (std. L/min)") + 
  theme_bw()

design <- "AABB
           CCDD"

p1 + p2 + p3 + p4 + plot_layout(design = design) + plot_annotation(tag_levels = "A")
  
```


## Figure 3

```{r sniff-volume, fig.height=7.5, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="A. Individual sniff volumes of individual canines, arranged from lightest to heaviest, left to right, and colored by body size (civilian canines) and operational status. B. Total sampling volumes for each trial by target chemical colored by trial and operational status. C. Volume flow rate for each trial by target chemical and colored by trial and operational status. On each plot, some outliers are not shown in order to show detail. Significance levels were designated * if p < 0.05, ** if p < 0.01, and *** if p < 0.001."}
# Figure 3 
trial_labels <- c("Sport Trial 1", "Sport Trial 2", "Sport Trial 3", "Law enforcement")
p1 <- ggplot(sniff_this, aes(x =  dog, y = sniff_volume, fill = size)) + 
  geom_point(pch = 21, alpha = 0.3, position = position_jitterdodge(jitter.width = 0.2)) + 
  geom_boxplot(na.rm = T, outlier.shape = NA, alpha = 0.6) +
  xlab("Participant number") + 
  ylab("Sniff volume (mL)")+
  ylim(0, 1000)+
  scale_fill_viridis(discrete = T, name = "", 
                    labels = c("Small (sport)", "Large (sport)", "Law\nenforcement"))+
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p3 <- ggplot(sniff_summaries_chemical, aes(x = chemical, y = total_volume, fill = trial)) + 
  geom_point(pch = 21, alpha = 0.3, position = position_jitterdodge(jitter.width = 0.2)) + 
  geom_boxplot(na.rm = T, outlier.shape = NA, alpha = 0.6) +
  ylab("Mean total sample\nvolume (mL)")+
  xlab(" ") +
  #ggtitle("Sniff volume by chemical")+
  scale_fill_manual(values = trial_cols, 
                    labels = trial_labels, 
                    name = " ") +
  scale_x_discrete(labels = chem_labels) +
  ylim(0, 8300) +
  #2e1h
  geom_segment(aes(x = 0.85, y = 7250, xend = 1.3, yend = 7250))+
  annotate("text", x = 1.1, y = 7500, label = "***")+
  #ammonia
  geom_segment(aes(x = 1.85, y = 7750, xend = 2.3, yend = 7750))+
  annotate("text", x = 2.1, y = 8000, label = "***")+
  #blank
  geom_segment(aes(x = 2.85, y = 7750, xend = 3.3, yend = 7750))+
  annotate("text", x = 3.1, y = 8000, label = "**")+
  #bromooctane
  geom_segment(aes(x = 3.85, y = 4000, xend = 4.3, yend = 4000))+
  annotate("text", x = 4.1, y = 4250, label = "***")+
  #methylbenzoate
  geom_segment(aes(x = 4.85, y = 6250, xend = 5.3, yend = 6250))+
  annotate("text", x = 5.1, y = 6500, label= "***")+
  
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p4 <- ggplot(sniff_results, aes(x = chemical, y = sniff_vfr, fill = trial)) + 
 geom_point(pch = 21, alpha = 0.3, position = position_jitterdodge(jitter.width = 0.2)) + 
  geom_boxplot(na.rm = T, outlier.shape = NA, alpha = 0.6) +
  ylab("Volume flow rate\n(mL/s)")+
  xlab(" ")+
  ylim(0, 15000)+
  scale_fill_manual(values = trial_cols, 
                    labels = trial_labels, 
                    name = " ") +
  scale_x_discrete(labels = chem_labels) +
  #ggtitle("Flow rate by chemical")+
  #2e1h
  geom_segment(aes(x = 0.85, y = 12000, xend = 1.3, yend = 12000))+
  annotate("text", x = 1.1, y = 12500, label = "***")+
  #ammonia
  geom_segment(aes(x = 1.85, y = 12000, xend = 2.3, yend = 12000))+
  annotate("text", x = 2.1, y = 12500, label = "***")+
  #blank
  geom_segment(aes(x = 2.85, y = 13500, xend = 3.3, yend = 13500))+
  annotate("text", x = 3.1, y = 14000, label = "**")+
  #bromooctane
  geom_segment(aes(x = 3.85, y = 13500, xend = 4.3, yend = 13500))+
  annotate("text", x = 4.1, y = 14000, label = "***")+
  #methylbenzoate
  geom_segment(aes(x = 4.85, y = 10000, xend = 5.3, yend = 10000))+
  annotate("text", x = 5.1, y = 10500, label= "***")+
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p1  / (p3 /p4 + plot_layout(guides = "collect") )  + plot_annotation(tag_levels = "A") +
  plot_layout(heights = c(1,3))
```

Accompanying statistical comparisons: 

```{r sniff-volume-fig-stats}
# Test Fig 3B panel: total sample volumes summarized by chemical
print("Test Fig 3B panel: total sample volumes summarized by chemical")
shapiro.test(sniff_summaries_chemical$total_volume) # data are non-normal

# 2E1H (civilian vs operational) SIG
snifftotvol_2E1H_model <- wilcox.test(
  sniff_summaries_chemical$total_volume[
    sniff_summaries_chemical$size == "operational" & 
      sniff_summaries_chemical$chemical == "2E1H"],
  sniff_summaries_chemical$total_volume[
    sniff_summaries_chemical$size != "operational" & 
      sniff_summaries_chemical$chemical == "2E1H"]) 
snifftotvol_2E1H_model$p.value <- p.adjust(snifftotvol_2E1H_model$p.value, 
                                        method = "bonferroni", n = 5)
print("Total volume: 2E1H sport vs law-enforcement")
print(snifftotvol_2E1H_model)

# Ammonia (civilian vs operational) SIG
snifftotvol_Ammonia_model <- wilcox.test(
  sniff_summaries_chemical$total_volume[
    sniff_summaries_chemical$size == "operational" & 
      sniff_summaries_chemical$chemical == "Ammonia"],                        
  sniff_summaries_chemical$total_volume[
    sniff_summaries_chemical$size != "operational" & 
      sniff_summaries_chemical$chemical == "Ammonia"])
snifftotvol_Ammonia_model$p.value <- p.adjust(snifftotvol_Ammonia_model$p.value, 
                                        method = "bonferroni", n = 5)
print("Total volume: Ammonia sport vs law-enforcement")
print(snifftotvol_Ammonia_model)

# Blank (civilian vs operational) SIG
snifftotvol_Blank_model <- wilcox.test(
  sniff_summaries_chemical$total_volume[
    sniff_summaries_chemical$size == "operational" &
      sniff_summaries_chemical$chemical == "Blank"],                        
  sniff_summaries_chemical$total_volume[
    sniff_summaries_chemical$size != "operational" & 
      sniff_summaries_chemical$chemical == "Blank"])
snifftotvol_Blank_model$p.value <- p.adjust(snifftotvol_Blank_model$p.value, 
                                        method = "bonferroni", n = 5)
print("Total volume: Blank sport vs law-enforcement")
print(snifftotvol_Blank_model)

# Bromooctane (civilian vs operational) SIG
snifftotvol_Bromo_model <- wilcox.test(
  sniff_summaries_chemical$total_volume[
    sniff_summaries_chemical$size == "operational" & 
      sniff_summaries_chemical$chemical == "bromooctane"],                        
  sniff_summaries_chemical$total_volume[
    sniff_summaries_chemical$size != "operational" & 
      sniff_summaries_chemical$chemical == "bromooctane"])
snifftotvol_Bromo_model$p.value <- p.adjust(snifftotvol_Bromo_model$p.value, 
                                        method = "bonferroni", n = 5)
print("Total volume: 1-BO sport vs law-enforcement")
print(snifftotvol_Bromo_model)


# Methyl benzoate (civilian vs operational) SIGNIFICANT
snifftotvol_Methyl_model <- wilcox.test(
  sniff_summaries_chemical$total_volume[
    sniff_summaries_chemical$size == "operational" &
    sniff_summaries_chemical$chemical == "methylbenzoate"],
  sniff_summaries_chemical$total_volume[
    sniff_summaries_chemical$size != "operational" & 
    sniff_summaries_chemical$chemical == "methylbenzoate"])
snifftotvol_Methyl_model$p.value <- p.adjust(snifftotvol_Methyl_model$p.value, 
                                        method = "bonferroni", n = 5)
print("Total volume: MB sport vs law-enforcement")
print(snifftotvol_Methyl_model)


#####################

# Test sniff-vfr p4 panel: sniff volume flow rate summarized by chemical
print("Test Volume flow rate Fig 3C: sniff volume flow rate summarized by chemical")
shapiro.test(sniff_summaries_chemical$sniff_vfr) # data are non-normal

# 2E1H (civilian vs operational) SIG
sniffvfr_2E1H_model <- wilcox.test(
  sniff_summaries_chemical$sniff_vfr[
    sniff_summaries_chemical$size == "operational" & 
    sniff_summaries_chemical$chemical == "2E1H"], 
  sniff_summaries_chemical$sniff_vfr[
    sniff_summaries_chemical$size != "operational" & 
    sniff_summaries_chemical$chemical == "2E1H"]) 
sniffvfr_2E1H_model$p.value <- p.adjust(sniffvfr_2E1H_model$p.value, 
                                        method = "bonferroni", n = 5)
print("Volume flow rate: 2E1H sport vs law-enforcement")
print(sniffvfr_2E1H_model)

# Ammonia (civilian vs operational) SIG
sniffvfr_Ammonia_model <- wilcox.test(
  sniff_summaries_chemical$sniff_vfr[
    sniff_summaries_chemical$size == "operational" & 
    sniff_summaries_chemical$chemical == "Ammonia"], 
  sniff_summaries_chemical$sniff_vfr[
    sniff_summaries_chemical$size != "operational" & 
    sniff_summaries_chemical$chemical == "Ammonia"]) 
sniffvfr_Ammonia_model$p.value <- p.adjust(sniffvfr_Ammonia_model$p.value, 
                                        method = "bonferroni", n = 5)
print("Volume flow rate: Ammonia sport vs law-enforcement")
print(sniffvfr_Ammonia_model)

# Blank (civilian vs operational) SIG
sniffvfr_Blank_model <- wilcox.test(
  sniff_summaries_chemical$sniff_vfr[
    sniff_summaries_chemical$size == "operational" & 
    sniff_summaries_chemical$chemical == "Blank"], 
  sniff_summaries_chemical$sniff_vfr[
    sniff_summaries_chemical$size != "operational" & 
    sniff_summaries_chemical$chemical == "Blank"]) 
sniffvfr_Blank_model$p.value <- p.adjust(sniffvfr_Blank_model$p.value, 
                                        method = "bonferroni", n = 5)
print("Volume flow rate: Blank sport vs law-enforcement")
print(sniffvfr_Blank_model)

# Bromooctane (civilian vs operational) SIG
sniffvfr_Bromo_model <- wilcox.test(
  sniff_summaries_chemical$sniff_vfr[
    sniff_summaries_chemical$size == "operational" & 
    sniff_summaries_chemical$chemical == "bromooctane"], 
  sniff_summaries_chemical$sniff_vfr[
    sniff_summaries_chemical$size != "operational" & 
    sniff_summaries_chemical$chemical == "bromooctane"]) 
sniffvfr_Bromo_model$p.value <- p.adjust(sniffvfr_Bromo_model$p.value, 
                                        method = "bonferroni", n = 5)
print("Volume flow rate: 1-BO sport vs law-enforcement")
print(sniffvfr_Bromo_model)

# Methyl benzoate (civilian vs operational) SIG
sniffvfr_Methyl_model <- wilcox.test(
  sniff_summaries_chemical$sniff_vfr[
    sniff_summaries_chemical$size == "operational" & 
    sniff_summaries_chemical$chemical == "methylbenzoate"], 
  sniff_summaries_chemical$sniff_vfr[
    sniff_summaries_chemical$size != "operational" & 
    sniff_summaries_chemical$chemical == "methylbenzoate"]) 
sniffvfr_Methyl_model$p.value <- p.adjust(sniffvfr_Methyl_model$p.value, 
                                        method = "bonferroni", n = 5)
print("Volume flow rate: MB sport vs law-enforcement")
print(sniffvfr_Methyl_model)
```

## Figure 4

```{r sniff-freq, fig.height=6.5, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="A. Sniff frequencies of all canines, arranged from heaviest to lightest, left to right. B: Sniff frequencies of each trial organized by chemical. Small canines were designated as less than 30 pounds. Significance levels were designated * if p < 0.05, ** if p < 0.01, and *** if p < 0.001."}
# Figure 6
p1 <- ggplot(sniff_results, aes(x = chemical, y = sniff_freq, fill = trial)) + 
  geom_point(pch = 21, alpha = 0.3, position = position_jitterdodge(jitter.width = 0.2)) + 
  geom_boxplot(na.rm = T, outlier.shape = NA, alpha = 0.6) +
  ylab("Sniff Frequency (sniff/s)")+
  xlab(" ")+
  ylim(0,10)+
  scale_fill_manual(values = trial_cols, 
                    labels = trial_labels, 
                    name = " ")+
  scale_x_discrete(labels = chem_labels) +
  geom_segment(aes(x = .65, y = 8.5, xend = 1.15, yend = 8.5))+
  annotate("text", x=.9, y=8.7, label= "*")+
  geom_segment(aes(x = 1.65, y = 8.5, xend = 2.15, yend = 8.5))+
  annotate("text", x=1.9, y=8.7, label= "**")+
  #ggtitle("Sniff Frequency by chemical")+
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p2 <- ggplot(sniff_results, aes(x = chemical, y = sniff_duration, fill = trial)) + 
  geom_point(pch = 21, alpha = 0.3, position = position_jitterdodge(jitter.width = 0.2)) + 
  geom_boxplot(na.rm = T, outlier.shape = NA, alpha = 0.6) +
  ylab("Sniff Duratoin (s)")+
  xlab(" ")+
  #ylim(0,10)+
  scale_fill_manual(values = trial_cols, 
                    labels = trial_labels, 
                    name = " ")+
  scale_x_discrete(labels = chem_labels) +
  #geom_segment(aes(x = .65, y = 8.5, xend = 1.15, yend = 8.5))+
  #annotate("text", x=.9, y=8.7, label= "*")+
  #geom_segment(aes(x = 1.65, y = 8.5, xend = 2.15, yend = 8.5))+
  #annotate("text", x=1.9, y=8.7, label= "**")+
  #ggtitle("Sniff Frequency by chemical")+
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p1 / p2 + plot_annotation(tag_levels = "A") + plot_layout(guides = "collect") 
```

Accompanying statistical comparisons: 

```{r sniff-vfr-stats}
# 2E1H (Trial 1 vs Trial 3) SIG
snifffreq_2E1H_model <- wilcox.test(
  sniff_summaries_chemical$sniff_freq[
    sniff_summaries_chemical$trial == 1 & 
    sniff_summaries_chemical$chemical == "2E1H"], 
  sniff_summaries_chemical$sniff_freq[
    sniff_summaries_chemical$trial == 3 & 
    sniff_summaries_chemical$chemical == "2E1H"]) 
print("Sniff frequency: 2E1H (Trial 1 vs Trial 3)")
print(snifffreq_2E1H_model)

# Ammonia (Trial 1 vs Trial 3) SIG
snifffreq_Amm_model <- wilcox.test(
  sniff_summaries_chemical$sniff_freq[
    sniff_summaries_chemical$trial == 1 & 
    sniff_summaries_chemical$chemical == "Ammonia"], 
  sniff_summaries_chemical$sniff_freq[
    sniff_summaries_chemical$trial == 3 & 
    sniff_summaries_chemical$chemical == "Ammonia"]) 
print("Sniff frequency: Ammonia (Trial 1 vs Trial 3)")
print(snifffreq_Amm_model)

# Blank (Trial 1 vs Trial 3) NOT
snifffreq_Blank_model <- wilcox.test(
  sniff_summaries_chemical$sniff_freq[
    sniff_summaries_chemical$trial == 1 & 
    sniff_summaries_chemical$chemical == "Blank"], 
  sniff_summaries_chemical$sniff_freq[
    sniff_summaries_chemical$trial == 3 & 
    sniff_summaries_chemical$chemical == "Blank"]) 
print("Sniff frequency: Blank (Trial 1 vs Trial 3)")
print(snifffreq_Blank_model)

# Bromooctane (Trial 1 vs Trial 3) NOT
snifffreq_Bromo_model <- wilcox.test(
  sniff_summaries_chemical$sniff_freq[
    sniff_summaries_chemical$trial == 1 & 
    sniff_summaries_chemical$chemical == "bromooctane"], 
  sniff_summaries_chemical$sniff_freq[
    sniff_summaries_chemical$trial == 3 & 
    sniff_summaries_chemical$chemical == "bromooctane"])
print("Sniff frequency: 1-BO (Trial 1 vs Trial 3)")
print(snifffreq_Bromo_model)

# Methyl benzoate (Trial 1 vs Trial 3) NOT
snifffreq_Methyl_model <- wilcox.test(
  sniff_summaries_chemical$sniff_freq[
    sniff_summaries_chemical$trial == 1 & 
    sniff_summaries_chemical$chemical == "methylbenzoate"], 
  sniff_summaries_chemical$sniff_freq[
    sniff_summaries_chemical$trial == 3 & 
    sniff_summaries_chemical$chemical == "methylbenzoate"]) 
print("Sniff frequency: Methyl benzoate (Trial 1 vs Trial 3)")
print(snifffreq_Methyl_model)
```

## Figure 5 

Figure reproduction: 

```{r volume-weight, fig.height=6.5, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="A. Mean sniff volume versus body weight colored by trial and operational status.  B. Mean sniff volume normalized by body weight versus snout length for civilian dogs. C. Mean sniff volume flow rate versus body weight colored by trial and operational status. D. Mean sniff volume flor rate versus snout length for civilian dogs. Some individual participants labeled by number. On each plot, linear regressions are indicated by blue lines."}

op_labels <- c("Sport", "Law enforcement")
fig4_dat <- sniff_summaries_alltrials
fig4_dat$labels <- fig4_dat$dog
#fig4_dat$labels <- ""
#id_labs <- c(1,2,3)
#fig4_dat$labels[id_labs] <- fig4_dat$dog[id_labs]

p1 <- ggplot(fig4_dat, 
             aes(weight, sniff_volume, fill = op, group = 1)) +
  geom_point(pch = 21, size = 2) +
  stat_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  geom_label_repel(aes(label = labels), fill = NA, max.overlaps = 2) +
  scale_fill_manual(values = trial_cols[c(1,4)], 
                    labels = op_labels, 
                    name = " ", drop = F) + 
  scale_color_manual(values = trial_cols[c(1,4)], 
                    labels = op_labels, 
                    name = " ", drop = F) + 
  xlab("Body mass (kg)") + 
  ylab("Mean sniff volume (mL)") +
  theme_bw()

p2 <- ggplot(na.omit(fig4_dat), aes(snout_length, norm_volume)) +
  geom_point(pch = 21, size = 2, color = "black", fill = trial_cols[1]) +
  stat_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  geom_label_repel(aes(label = labels), fill = NA, max.overlaps = 3) +
  #scale_fill_manual(values = trial_cols, 
  #                  labels = trial_labels, 
  #                  name = " ", drop = F) + 
  #scale_color_manual(values = trial_cols, 
  #                  labels = trial_labels, 
  #                  name = " ", drop = F) +
  xlab("Snout length (cm)") + 
  ylab("Normalized sniff volume (mL/kg)") +
  theme_bw()

p3 <- ggplot(fig4_dat, 
             aes(weight, sniff_vfr, fill = op, group = 1)) +
  geom_point(pch = 21, size = 2) +
  stat_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  geom_label_repel(aes(label = labels), fill = NA, max.overlaps = 3) +
  scale_fill_manual(values = trial_cols[c(1,4)], 
                    labels = op_labels, 
                    name = " ", drop = F) + 
  scale_color_manual(values = trial_cols, 
                    labels = op_labels, 
                    name = " ", drop = F) + 
  xlab("Body mass (kg)") + 
  ylab("Mean sniff volume flow rate (mL/s)") +
  theme_bw()

p4 <- ggplot(na.omit(fig4_dat), aes(snout_length, sniff_vfr/weight)) +
  geom_point(pch = 21, size = 2, color = "black", fill = trial_cols[1]) +
  stat_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  geom_label_repel(aes(label = labels), fill = NA, max.overlaps = 3) +
  #scale_fill_manual(values = trial_cols, 
  #                  labels = trial_labels, 
  #                  name = " ", drop = F) + 
  #scale_color_manual(values = trial_cols, 
  #                  labels = trial_labels, 
  #                  name = " ", drop = F) +
  xlab("Snout length (cm)") + 
  ylab("Normalized sniff volume\nflow rate (mL/s*kg)") +
  theme_bw()

(p1 + p2) / (p3 + p4 ) + plot_annotation(tag_levels = "A") + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```


Accompanying statistical analyses: 

```{r fig-5-stats}
## Linear regressions accompaning panels on Fig. 5
sport_reg_Fig5A <- summary(lm(sniff_volume~weight, 
                              data = sniff_summaries_alltrials[sniff_summaries_alltrials$op == "civilian",]))
print("Regression Fig 5A: sport only")
print(sport_reg_Fig5A)

all_reg_Fig5A <- summary(lm(sniff_volume~weight, 
                              data = sniff_summaries_alltrials))
print("Regression Fig 5A: all dogs")
print(all_reg_Fig5A)

sport_reg_Fig5B <- summary(lm(norm_volume~snout_length, data = sniff_summaries_alltrials))
print("Regression Fig 5B")
print(sport_reg_Fig5B)
```
# Reproduction of statistical values in text of manuscript


```{r text-stats}

## Comparing sniff volumes between operational and civilian dogs for all trials

# Testing data for normality:
# shapiro.test(sniff_summaries_alltrials$sniff_volume)
# entire set is non-normal
# shapiro.test(sniff_summaries_alltrials$sniff_volume[sniff_summaries_alltrials$op == "civilian"])
# data are non-normal
# shapiro.test(sniff_summaries_alltrials$sniff_volume[sniff_summaries_alltrials$op == "operational"])
# data are normal

# Performing non-parametric comparison:
comp_op_volume_all <- wilcox.test(sniff_summaries_alltrials$sniff_volume[sniff_summaries_alltrials$op == 
                                                                    "civilian"],
                           sniff_summaries_alltrials$sniff_volume[sniff_summaries_alltrials$op ==
                                                                    "operational"])
print("Comparing sniff volumes between operational and civilian dogs for all trials")
print(comp_op_volume_all)

## Comparing sniff volume flow rates between operational and civilian dogs for all trials

# Testing data for normality:
# shapiro.test(sniff_summaries_alltrials$sniff_vfr)
# entire set is non-normal
# shapiro.test(sniff_summaries_alltrials$sniff_vfr[sniff_summaries_alltrials$op == "civilian"])
# data are non-normal
# shapiro.test(sniff_summaries_alltrials$sniff_vfr[sniff_summaries_alltrials$op == "operational"])
# data are normal

# Performing non-parametric comparison:
comp_op_vfr_all <- wilcox.test(sniff_summaries_alltrials$sniff_vfr[sniff_summaries_alltrials$op == 
                                                                    "civilian"],
                           sniff_summaries_alltrials$sniff_vfr[sniff_summaries_alltrials$op ==
                                                                    "operational"])
print("Comparing sniff volume flow rates between operational and civilian dogs for all trials")
print(comp_op_vfr_all)
## Comparing sniffing frequencies between trials 1 and 3 for target chemicals only

# Testing data for normality
# shapiro.test(sniff_summaries_chemical$sniff_freq[sniff_summaries_chemical$trial == 1 & 
#                                                   sniff_summaries_chemical$target == "yes"])
# data are non-normal
# shapiro.test(sniff_summaries_chemical$sniff_freq[sniff_summaries_chemical$trial == 3 & 
#                                                   sniff_summaries_chemical$target == "yes"])
# data are normal
#shapiro.test(c(sniff_summaries_chemical$sniff_freq[sniff_summaries_chemical$trial == 1 & 
#                                                   sniff_summaries_chemical$target == "yes"], 
#            sniff_summaries_chemical$sniff_freq[sniff_summaries_chemical$trial == 3 & 
#                                                   sniff_summaries_chemical$target == "yes"]))
# entire data set is normal

# Performing non-parametric comparison:
comp_target_trial13 <- wilcox.test(sniff_summaries_chemical$sniff_freq[sniff_summaries_chemical$trial == 1 & 
                                                   sniff_summaries_chemical$target == "yes"], 
            sniff_summaries_chemical$sniff_freq[sniff_summaries_chemical$trial == 3 & 
                                                   sniff_summaries_chemical$target == "yes"])
print(comp_target_trial13)
# Similar result if t-test is used:
#t.test(sniff_summaries_chemical$sniff_freq[sniff_summaries_chemical$trial == 1 & 
#                                                   sniff_summaries_chemical$target == "yes"], 
#            sniff_summaries_chemical$sniff_freq[sniff_summaries_chemical$trial == 3 & 
#                                                   sniff_summaries_chemical$target == "yes"])
```